Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-01-12 analysis/paper/paper.Rmd
moved creation of flwr and next year columns to demography-wrangling.r.  Also changed how dead plants are calculated (closes #2)
worked on statistical methods

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig. -- <!--# fig ref -->). These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried. The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al. 2001) <!--# replace ref -->. The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al. 1992) <!--# replace ref -->, and an understory dominated by stemless palms (Scariot 1999). The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H. acuminata*---are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X <!--# from data? --> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

Monthly precipitation and potential evapotranspiration records ( <!--# date range -->) were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. We calculated 3-month SPEI using the `SPEI` package in \`r version\$version.string\` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. <!--# something about why SPEI-3 is appropriate. --> Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->. We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).

@@

Growth

Survival

Flowering probability

>>>>

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig. -- <!--# fig ref -->). These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried. The vegetation regenerating around fragments is cleared periodically to ensure their isolation [@bierregaardLessonsAmazoniaEcology2001]. The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999]. The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H. acuminata*---are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants will have 1-X <!--# from data? --> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method. Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months. SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckee1993]. We chose a 3 month time scale as most relevant for *H. acuminata* as plants do not access sources of water buffered over longer time scales such as ground water or flood water <!--# work on this sentence -->. We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database <!--# citation --> and TRMM <!--# citation --> and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1)<!--# ref to appendix 1 -->.

@@

To model growth, we used shoot number as a proxy for plant size. Shoot number is easy to measure accurately and, unlike plant height, does not change if a plant is knocked over by a treefall. <!--# I could consider an integrative measure like log(shts * ht) -->. Size in year $t+1$ was modeled as a function of the natural log of size in year $t$ and drought history (i.e. the crossbasis function) with plot as a random effect. We used a scaled t error family distribution with a log link since residuals were leptokurtic when a gaussian distribution was used.

For survival, plants were assumed dead the year after they were last observed if the plant was not observed for three or more years. Survival and flowering probability were similarly modeled as a function of size and drought history (i.e. the crossbasis function) with a random effect of plot, but with a binomial family error distribution and a logit link function.

Statistical significance of crossbasis functions and covariates was determined by ....

To better understand the shape of the crossbasis function, we calculated fitted values for an average sized plant given a range of possible SPEI values from \_\_\_ to \_\_\_ at different lag times from \_\_\_ to \_\_\_ months, holding all other SPEI values at the average across the dataset (\_\_\_). These results could then be plotted as heatmaps with the *ggplot2* package ( <!--# cite ggplot2 --> ).

One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function. This limitation meant that we were unable to model a potential size by environment (i.e. drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018]. To check for a size by environment interaction, we followed the methods of Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package <!--# cite lme4 -->. Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year. For each vital rate, we included an additional random effect of year on either the intercept or both the intercept and slope of the relationship between plant size and the response. We tested the significance of the year by size interaction with a permutation <!--# not quite permutation, I think?  Check trednnick et al. again for what they call it --> test, again following Treddnick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018].
